---
title: "Practical Machine Learning Prediction Assignment"
author: "Karishma Pandey"
output: html_document
---

# Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit, it is
possible to collect large amounts of data about personal activity.
This project uses accelerometer measurements collected from the
belt, forearm, arm, and dumbbell of six participants performing
barbell lifts in five different ways.

The objective of this project is to predict the manner in which the
exercise was performed. The response variable is `classe`, with
five possible classes: A, B, C, D, and E.

# Data

The training dataset contains observations collected from
accelerometers attached to different parts of the participants'
bodies. The testing dataset contains 20 observations for which
the exercise classification must be predicted.

The training and testing datasets were downloaded from the
sources provided in the course project instructions.

# Loading Required Libraries

```{r setup, message=FALSE, warning=FALSE}
library(caret)
library(randomForest)

set.seed(12345)
```

# Loading the Data

```{r load-data}
training <- read.csv(
  "pml-training.csv",
  na.strings = c("", "NA", "#DIV/0!")
)

testing <- read.csv(
  "pml-testing.csv",
  na.strings = c("", "NA", "#DIV/0!")
)

dim(training)
dim(testing)
```

# Exploratory Data Analysis

The variable `classe` is the outcome variable that we want to
predict. It represents the manner in which the exercise was
performed.

```{r class-distribution}
table(training$classe)
```

# Data Cleaning

The dataset contains several variables with a large number of
missing values. Variables containing more than 60% missing values
are removed. Near-zero variance variables are also removed because
they provide little useful information for classification.

The outcome variable `classe` is excluded from the predictors.
Identification and time-related variables are also removed.

```{r data-cleaning}

# Separate the outcome variable from the predictors
training_predictors <- training[, setdiff(names(training), "classe")]

# Calculate the proportion of missing values in each predictor
missing_fraction <- sapply(
  training_predictors,
  function(x) mean(is.na(x))
)

# Keep predictors with less than 60% missing values
keep_columns <- names(
  missing_fraction[missing_fraction < 0.60]
)

# Make sure only variables available in both datasets are selected
keep_columns <- intersect(
  keep_columns,
  names(testing)
)

# Create cleaned datasets
training_clean <- training[, c(keep_columns, "classe")]
testing_clean <- testing[, keep_columns]

# Remove near-zero variance predictors
nzv <- nearZeroVar(training_clean[, keep_columns])

if (length(nzv) > 0) {
  keep_columns <- keep_columns[-nzv]
}

# Recreate cleaned datasets after NZV removal
training_clean <- training[, c(keep_columns, "classe")]
testing_clean <- testing[, keep_columns]

# Remove the first seven identification/time-related variables
training_clean <- training_clean[, -c(1:7)]
testing_clean <- testing_clean[, -c(1:7)]

# Check dimensions
dim(training_clean)
dim(testing_clean)
```

```{r missing-check}
sum(is.na(training_clean))
sum(is.na(testing_clean))
```

# Training and Validation Split

The cleaned training data are divided into a training set and a
validation set. Approximately 70% of the observations are used
for model training and the remaining 30% are reserved for
validation. The validation data are not used to fit the model,
allowing its performance to be evaluated on previously unseen
observations.

```{r split-data}

set.seed(12345)

train_index <- createDataPartition(
  training_clean$classe,
  p = 0.70,
  list = FALSE
)

train_data <- training_clean[train_index, ]

validation_data <- training_clean[-train_index, ]

dim(train_data)
dim(validation_data)
```

# Random Forest Model

A Random Forest classifier is used because the response variable
`classe` is categorical and the dataset contains multiple
sensor-based predictor variables. Five-fold cross-validation is
used during model training to assess model performance across
different subsets of the training data.

```{r random-forest, message=FALSE, warning=FALSE}

set.seed(12345)

control <- trainControl(
  method = "cv",
  number = 5
)

rf_model <- train(
  classe ~ .,
  data = train_data,
  method = "rf",
  trControl = control,
  ntree = 300
)

rf_model
```

# Model Evaluation

The trained Random Forest model is evaluated on the validation
dataset, which was not used during model training.

```{r validation-results}

validation_predictions <- predict(
  rf_model,
  newdata = validation_data
)

# Ensure predictions and reference values have identical factor levels
validation_predictions <- factor(
  validation_predictions,
  levels = levels(factor(training_clean$classe))
)

validation_actual <- factor(
  validation_data$classe,
  levels = levels(factor(training_clean$classe))
)

confusionMatrix(
  validation_predictions,
  validation_actual
)
```

## Expected Out-of-Sample Error

The validation accuracy is used to estimate the expected
out-of-sample error. The error rate is calculated as one minus
the validation accuracy.

```{r out-of-sample-error}

accuracy <- confusionMatrix(
  validation_predictions,
  validation_actual
)$overall["Accuracy"]

out_of_sample_error <- 1 - accuracy

accuracy
out_of_sample_error
```

# Variable Importance

Variable importance is examined to identify the predictor variables
that contributed most to the Random Forest classification.

```{r variable-importance, fig.width=8, fig.height=6}

plot(
  varImp(rf_model),
  top = 15,
  main = "Top 15 Predictor Variables"
)
```

# Final Model

After evaluating the Random Forest model on the validation data,
the final model is trained using the complete cleaned training
dataset. This allows the model to use all available labeled
observations before making predictions on the 20 test cases.

```{r final-model, message=FALSE, warning=FALSE}

set.seed(12345)

final_model <- train(
  classe ~ .,
  data = training_clean,
  method = "rf",
  trControl = trainControl(method = "none"),
  tuneGrid = data.frame(mtry = 26),
  ntree = 300
)

final_model
```

# Prediction of the Test Cases

The final Random Forest model is used to predict the exercise
classification for the 20 observations in the provided testing
dataset.

```{r test-predictions}

final_predictions <- predict(
  final_model,
  newdata = testing_clean
)

final_predictions
```

```{r prediction-list}

prediction_table <- data.frame(
  Problem_ID = 1:20,
  Prediction = final_predictions
)

prediction_table
```

# Conclusion

A Random Forest classification model was developed to predict
the manner in which participants performed the weight lifting
exercise. The data were cleaned by removing variables with
excessive missing values, near-zero variance variables, and
identification-related variables.

The model was trained using five-fold cross-validation. The
best model used an `mtry` value of 26 and achieved a validation
accuracy of approximately 99.47%. The corresponding estimated
out-of-sample error was approximately 0.53%.

The final Random Forest model was then trained using the complete
cleaned training dataset. This final model was used to predict
the exercise classification for the 20 observations in the
provided testing dataset.

